---
title: "Crossjoins"
description: |
  Exploring crossjoins to find all possible combinations of elements in a list
author:
  - name: Jolene
date: 09-08-2021
output:
  distill::distill_article:
    self_contained: false
draft: true
---

Suppose we want to find all combinations of desserts from a list of fruits and a list of dessert types. `{data.table}` has a crossjoin function that is extremely useful to achieve that. `CJ` automatically arranges the vectors alphabetically.

```{r setup, include=TRUE}
library(purrr)
library(crossmap)
library(data.table)
library(rbenchmark)
list <- list(
  fruit <- c('apple','apple', 'orange', 'pear', 'peach'), 
  type <- c('cake', 'tart', 'pie'),
  s <- c('s', 'd', 'f')
)

# CJ(x,y) forms a data.table from the cross product of the vectors 
CJ(fruit, type, unique = T)

```

Sidenote: I found that if I changed all the assignment operators for the list to equals, `fruit` and `type` become vectors within the list but do not exist outside the list. In other words, if I try to view `fruit`, R returns `Error: object 'fruit' not found`. 

Since `CJ(x,y)` is a `data.table`, the `{data.table}` syntax `DT[...]` can be used.

```{r}
dessert_combi_dt <- CJ(fruit, type)[, dessert := paste(fruit, type)]  
dessert_combi_dt
```

Or, if you just require a vector of the combinations 

```{r}
dessert_combi <- CJ(fruit, type)[,paste(fruit, type)]  
dessert_combi
```

This post was inspired by a tip I saw online, using `xmap_` functions from `{crossmap}` `r emo::ji('package')`